---
title: "Model"
date: "Last knitted on `r format(Sys.Date(), '%d %b %Y')`"
author: "Leavitt, Kopf, Weber, Pearson"
output:
  html_document: 
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: true
    toc_depth: 3
    code_folding: show
editor_options:
  chunk_output_type: console # switch to 'inline' if preferred
---

# Setup

```{r "setup", message = FALSE}
# packages
library(tidyverse) # general data wrangling and plotting
library(nlstools)

# scripts
source("scripts/plotting_functions.R")
source("scripts/error_propagation.R")

# global knitting options for code rendering
knitr::opts_chunk$set(
  eval = TRUE, echo = FALSE,
  collapse = TRUE, comment = "#>",
  dev = c("png", "pdf", "postscript"),
  dev.args = list(pdf = list(encoding = "WinAnsi", useDingbats = FALSE)),
  fig.keep = "all",
  fig.path = file.path("plots", "")
)
```

> NOTE: please knit the `01_calculations.Rmd` notebook first to prepare the data used for plotting.

# H accounting

$$
\begin{aligned}
n &= 80 - 2 \cdot r \\
f_{S} &= \frac{24 + (23 + 1/3) \cdot (1 - x) - r}{n} \\
f_{LS_W} &= \frac{2/3 \cdot (1 - x)}{n} \\
f_{RS_W} &=  \frac{r}{n} \\
f_{GGR_W} &= \frac{8 - r}{n} \\
f_{LS_{NADPH}} &= \frac{16}{n} \\
f_{GGR_{FAD}} &= \frac{8 - r}{n} \\
f_{x} &= \frac{24 \cdot x}{n} 
\end{aligned}
$$


## N. maritimus

With $H_{Act} = 2/3 \cdot H_{H2O} + 1/3 \cdot H_{NADPH}$ for *N. maritimus*:

$$
\begin{aligned}
f_{*A} &= 0 \\
f_{*W} &= 2/3 \cdot f_{A_W} + f_{LS_W} + f_{GGR_W} + f_{RS_W} \\
 &= \frac{16 + (15 + 5/9) \cdot (1 - x) - 2/3 \cdot r}{n} + \frac{2/3 \cdot (1 - x)}{n} + \frac{8 - r}{n} + \frac{r}{n} \\
 & = \frac{24 + 146/9 \cdot (1 - x) - 2/3 \cdot r}{80 - 2 \cdot r} \\
f_{*NADPH} &= 1/3 \cdot f_{A} + f_{LS_{NADPH}} \\
  &= \frac{8 + (7 + 7/9) \cdot (1 - x) - 1/3 \cdot r}{n} + \frac{16}{n} \\
  &= \frac{24 + 70/9 \cdot (1 - x) - 1/3 \cdot r}{80 - 2 \cdot r} \\
\end{aligned}
$$

## Functions

```{r}
# number of hydrogens
n_BP <- function(nr) 80 - 2 * nr
n_A <- function(x, nr) 24 + (23 + 1/3) * (1 - x) - nr
n_LS_W <- function(x) 2/3 * (1 - x)
n_GGR_W <- function(nr) 8 - nr
n_RS_W <- function(nr) nr
n_A_W <- function(x, nr) 2/3 * n_A(x, nr)
n_LS_NADPH <- function() 16
n_A_NADPH <- function (x, nr) 1/3 * n_A(x, nr)
n_FAD <- function(nr) 8 - nr
n_x <- function(x) 24 * x

f_W <- function(nr, x) (n_LS_W(x) + n_GGR_W(nr) + n_RS_W(nr) + n_A_W(x, nr)) / n_BP(nr)
f_NADPH <- function(nr, x) (n_LS_NADPH() + n_A_NADPH(x, nr)) / n_BP(nr)
f_FAD <- function(nr, x) n_FAD(nr) / n_BP(nr)
f_x <- function(nr, x) n_x(x) / n_BP(nr)
```

## Results

```{r}
# H accounting calculations
H_accounting <- 
  tibble(
    # FIXME: scenarios but with GGR H instead of FD
    scenario = c("Scenario 2", "Scenario 4"),
    x = c(0, 1)
  ) %>% 
  crossing(select(readxl::read_excel("data/compounds.xlsx"), bp_short, n_rings)) %>%
  mutate(
    n_BP = n_BP(n_rings),
    n_A_W = n_A_W(x, n_rings),
    n_LS_W = n_LS_W(x),
    n_GGR_W = n_GGR_W(n_rings),
    n_RS_W = n_RS_W(n_rings),
    n_W = n_A_W + n_LS_W + n_GGR_W + n_RS_W,
    n_A_NADPH = n_A_NADPH(x, n_rings),
    n_LS_NADPH = n_LS_NADPH(),
    n_NADPH = n_A_NADPH + n_LS_NADPH,
    n_FAD = n_FAD(n_rings),
    n_x = n_x(x),
    check = near(n_A_W + n_LS_W + n_GGR_W + n_RS_W + n_A_NADPH + n_LS_NADPH + n_FAD + n_x, n_BP)
  )

# save in cache
write_rds(H_accounting, "cache/H_accounting.rds")
```


# Model

## BP Components

$$
\begin{aligned}
\text{all }&\text{biphytane producers:} \\
BP &= f_A + (f_{LS_W} + f_{GGR_W} + f_{RS_W}) + f_{LS_{NADPH}} + f_{GGR_{FAD}} + f_x \\
 &= f_A + f_W + f_{NADPH} + f_{FAD} +f_x \\
 \text{N. }&\text{maritimus:} \\
 BP &= (f_{A_W} + f_{LS_W} + f_{GGR_W} + f_{RS_W}) + (f_{A_{NADPH}} + f_{LS_{NADPH}}) + f_{GGR_{FAD}} + f_x \\
 &= f_{*W} + f_{*NADPH} + f_{FAD} +f_x \\
\end{aligned}
$$

## Elecron flux

Parametrizing $f_L$ (NADH loss terms) with doubling time:

$$
\begin{aligned}
f_L &= \frac{\phi_{L}}{\phi_{ETC}} = 1 - \frac{\phi_{TH}}{\phi_{ETC}} = 1 - f_N \\
f_L &= \lambda \cdot (1 - X_{TD}) \\
\text{B} &= \frac{- \phi_{NADH\rightarrow NADPH}}{\phi_{ETC}} = -f_N = f_L - 1
\end{aligned}
$$

## R_NADPH

$$
\begin{aligned}
R_{NADH} &= 
  \frac{\alpha_{ETC} \cdot R_W}{\alpha_{TH} + f_L \cdot (1 - \alpha_{TH})} \\
  &= \frac{\alpha_{ETC} \cdot R_W}{\alpha_{TH} + \lambda \cdot (1 - X_{TD}) \cdot (1 - \alpha_{TH})} \\
R_{NADPH} &= \alpha_{TH} \cdot R_{NADH} \\
  &= \frac{\alpha_{TH} \cdot \alpha_{ETC} \cdot R_W}{\alpha_{TH} + \lambda \cdot (1 - X_{TD}) \cdot (1 - \alpha_{TH})} \\

\end{aligned}
$$

## R_BP (N. maritimus)

$$
\begin{aligned}
R_{BP} &= \left( \alpha_W   f_{*W} + \alpha_x  f_x \right) R_W + \left(\alpha_{NADPH} f_{*NADPH} + \alpha_{FAD} f_{FAD} \right) R_{NADPH} \\
 &= \left[\alpha_W   f_{*W} + \alpha_x  f_x + 
  \left( \alpha_{NADPH} f_{*NADPH} + \alpha_{FAD} f_{FAD} \right) \frac{\alpha_{ETC} \alpha_{TH}}
  {\alpha_{TH} + \lambda  (1 - X_{TD}) (1 - \alpha_{TH})}  \right] R_W \\
\rightarrow \alpha_{BP/W} &= \frac{R_{BP}}{R_W} \\
  &= \alpha_W   f_{*W} + \alpha_x  f_x + 
  \alpha_{ETC} \cdot \left( \alpha_{NADPH} f_{*NADPH} + \alpha_{FAD} f_{FAD} \right) \frac{1}
  {1 + \lambda  (1 - X_{TD}) (1/\alpha_{TH} - 1)} \\
  &= \alpha_W   f_{*W} + \alpha_x  f_x +\alpha_{ETC} \cdot \left( \alpha_{NADPH} f_{*NADPH} + \alpha_{FAD} f_{FAD} \right) \cdot \frac{1}{1 + k \cdot (1 - X_{TD})}\\
\text{with } &k = \frac{\lambda}{\alpha_{TH}} - \lambda 
  \\ \rightarrow &\lambda = \frac{\alpha_{TH}}{1 - \alpha_{TH}} \cdot k
  \\ \rightarrow &\alpha_{TH} = \frac{\lambda}{k + \lambda}
\end{aligned}
$$


# Fit

## Data

```{r}
TDmin <- 20
TDmax <- 120
calculate_XTD <- function(TD.hours, min = TDmin, max = TDmax) {
  return((TD.hours - min) / (max - min))
}

compounds <- readxl::read_excel("data/compounds.xlsx")
metadata <- readxl::read_excel("data/metadata.xlsx")
model_data <- readr::read_rds("cache/samples_sum.rds") %>%
  left_join(compounds, by = "compound") %>%
  left_join(metadata, by = c("organism", "growth")) %>%
  select(bp_short, n_rings, eps = eps_weighted_mean.permil, growth, TD.hours) %>%
  mutate(
    alpha = eps/1000 + 1,
    XTD = calculate_XTD(TD.hours)
  )

model_data
```

## NLS

```{r}
# calculate model output
calculate_model_output <- function(n_rings, XTD, alpha_W, x, alpha_NADPH, alpha_FAD_vs_NADPH, k) {
  alpha_ETC <- 1
  alpha_x <- alpha_W
  f_W <- f_W(n_rings, x)
  f_NADPH <- f_NADPH(n_rings, x)
  f_FAD <- f_FAD(n_rings, x)
  f_x <- f_x(n_rings, x)
  alpha_W * f_W + alpha_x * f_x + alpha_ETC * alpha_NADPH * (f_NADPH + alpha_FAD_vs_NADPH * f_FAD) / (1 + k * (1 - XTD))
}

# non-linear least squares fit of model
#start_values <- list(alpha_NADPH = 0.1, alpha_FAD_vs_NADPH = 0.2, k = 0.1)
set.seed(123)
start_values <- list(alpha_NADPH = runif(1), alpha_FAD_vs_NADPH = runif(1), k = runif(1))
message("Fitting with random start values: ")
print(start_values)
fit_model_by_nls <- function(alpha_W, x) {
  nls(
    ~(calculate_model_output(n_rings, XTD, alpha_W, x, alpha_NADPH, alpha_FAD_vs_NADPH, k) - alpha) * 1000, 
    data = mutate(model_data, alpha_W = !!alpha_W, x = !!x),
    start = start_values
  )
}

# run model for alpha_W and x combinations
fits <- 
  crossing(
    alpha_W = c(0.5, 0.7, 0.9),
    x = c(0, 1)
  ) %>% 
  mutate(
    nls_fit = pmap(list(alpha_W, x), fit_model_by_nls),
    nls_sum = map(nls_fit, broom::glance),
    nls_sigma = map_dbl(nls_sum, ~.x$sigma),
    nls_coefs = map(nls_fit, ~{
      left_join(
        broom::tidy(.x),
        nlstools::confint2(.x, level = 0.95, method = "asymptotic") %>% 
          as_tibble(rownames = "term"),
        by = "term"
      )
    })
  ) %>%
  select(-nls_fit, -nls_sum) %>%
  unnest(nls_coefs)
```


# Results

```{r}
# pivot wide
fits_wide <- fits %>%
  select(-statistic, -p.value) %>%
  pivot_wider(
    names_from = term,
    values_from = c(estimate, std.error, `2.5 %`, `97.5 %`)
  ) %>%
  transmute(
    resid.permil = nls_sigma,
    x, 
    x_fct = as_factor(x) %>% fct_recode(
      "none" = "0",
      "max" = "1"
    ),
    alpha_W = alpha_W,
    alpha_NADPH = estimate_alpha_NADPH,
    alpha_NADPH_se = std.error_alpha_NADPH,
    alpha_NADPH_ci_2.5 = `2.5 %_alpha_NADPH`,
    alpha_NADPH_ci_97.5 = `97.5 %_alpha_NADPH`,
    alpha_FAD_vs_NADPH = estimate_alpha_FAD_vs_NADPH,
    alpha_FAD_vs_NADPH_se = std.error_alpha_FAD_vs_NADPH,
    alpha_FAD_vs_NADPH_ci_2.5 = `2.5 %_alpha_FAD_vs_NADPH`,
    alpha_FAD_vs_NADPH_ci_97.5 = `97.5 %_alpha_FAD_vs_NADPH`,
    k = estimate_k,
    k_se = std.error_k
  ) 
head(fits_wide)

# alpha_TH and lambda calculated from k
calculate_lambda_expr <- rlang::expr(alpha_TH / (1 - alpha_TH) * k)
alpha_TH_lims <- c(0.222, 0.566) # literature constraints
fits_wide_w_lambdas <-
  fits_wide %>%
  crossing(alpha_TH = seq(0.08, 0.99, by = 0.01)) %>%
  mutate(
    lambda = eval(calculate_lambda_expr),
    lambda_se = propagate_error(!!calculate_lambda_expr, dalpha_TH = 0, dk = k_se)
  )

# save in cache
write_rds(alpha_TH_lims, "cache/alpha_TH_lims.rds")
write_rds(fits_wide, "cache/fits_wide.rds")
write_rds(fits_wide_w_lambdas, "cache/fits_wide_w_lambdas.rds")
```



## Alphas from model fits

```{r}
# model points (TD of original data point + min and max)
model_points <- 
  bind_rows(
    model_data %>% select(bp_short, n_rings, growth, TD.hours, alpha),
    model_data %>%
      select(bp_short, growth, n_rings) %>% unique() %>%
      crossing(TD.hours = c(TDmin, TDmax))
  ) %>%
  arrange(bp_short, TD.hours)

# calculate alpha values from model fits
fits_alphas <- 
  fits_wide %>%
  mutate(model_id = row_number()) %>%
  crossing(model_points) %>%
  mutate(
    XTD = calculate_XTD(TD.hours),
    m_alpha = calculate_model_output(
      n_rings, XTD, alpha_W, x, alpha_NADPH, alpha_FAD_vs_NADPH, k
    )
  ) 

# standard deviation of residuals - recalculation to check
fits_alphas %>% 
  filter(!is.na(alpha)) %>%
  group_by(model_id, x, x_fct, alpha_W, alpha_NADPH, alpha_FAD_vs_NADPH, k, resid.permil) %>%
  summarize(
    # degrees of freedom based on the 3 parameter fits
    resid_recalc.permil = sqrt(sum((m_alpha * 1000 - alpha * 1000)^2) / (n() - 3)),
    .groups = "drop"
  )

# averages across all model fits
fits_alphas_sum <-
  fits_alphas %>%
  group_by(bp_short, n_rings, growth, TD.hours) %>%
  summarize(
    m_alpha_sd = sd(m_alpha),
    m_alpha = mean(m_alpha),
    .groups = "drop"
  ) %>%
  arrange(growth, TD.hours, bp_short)

# save in cache
write_rds(fits_alphas_sum, "cache/fits_alphas_sum.rds")
```

## Ring enrichment

```{r}
# calculate model ring enrichments
fits_ring_enrichments <-
  fits_alphas %>%
  select(model_id, n_rings, TD.hours, m_alpha) %>%
  { full_join(., ., by = c("model_id", "TD.hours"))  } %>%
  filter(n_rings.x > n_rings.y) %>%
  mutate(
   D_eps_per_ring = 1000 * (m_alpha.y - m_alpha.x)/(n_rings.y - n_rings.x)
  ) %>%
  group_by(n_rings.x) %>%
  summarize(
    D_eps_per_ring_sd = sd(D_eps_per_ring),
    D_eps_per_ring_median = median(D_eps_per_ring),
    D_eps_per_ring = mean(D_eps_per_ring),
    .groups = "drop"
  )

# overall
fits_ring_enrichments %>%
  summarize(
    permil_per_ring_mean = mean(D_eps_per_ring),
    permil_per_ring_median = median(D_eps_per_ring),
    permil_per_ring_sd = sd(D_eps_per_ring)
  )
```

## Growth regression

```{r}
# calculate model growth regressions
fits_growth_regs <- 
  fits_alphas_sum %>%
  nest(data = -c(bp_short, n_rings)) %>%
  mutate(
    fit = map(data, ~lm(m_alpha ~ TD.hours, data = .x)),
    estimates = map(fit, broom::tidy)
  ) %>%
  unnest(estimates) %>% 
  filter(term == "TD.hours") %>%
  select(-fit)

# overall
fits_growth_regs %>%
  summarize(
    permil_per_hour_mean = 1000 * mean(estimate),
    permil_per_hour_median= 1000 * median(estimate),
    permil_per_hour_sd = 1000 * sd(estimate)
  )
```

## NADPH imbalance

```{r}
# calculate NADPh imbalance based on the alpha_TH mid value
NADPH_imbalances <- 
  fits_wide_w_lambdas %>%
  crossing(metadata) %>%
  mutate(
    XTD = calculate_XTD(TD.hours),
    fL = lambda * (1 - XTD),
    NADPH_imb = fL - 1
  )

# save in cache
write_rds(NADPH_imbalances, "cache/NADPH_imbalances.rds")
```


# DRAFT literature data

$$
\begin{aligned}
R_{NADPH} &= \alpha_{ETC} \cdot R_W \\
R_{BP} &= \left( \alpha_W   f_{W} + \alpha_x  f_x \right) R_W + \left(\alpha_{NADPH} f_{NADPH} + \alpha_{FAD} f_{FAD} \right) R_{NADPH} \\
\rightarrow \alpha_{BP/W} &= \alpha_W   f_{W} + \alpha_x  f_x + 
  \left( \alpha_{NADPH} f_{NADPH} + \alpha_{FAD} f_{FAD} \right) \alpha_{ETC}

\end{aligned}
$$

# TODO: fit these data with variable fw and fNADPH (fmn somewhat fixed, NADPH assuming minimum form MVA pathway, fx = 0). 

